\documentclass{article}

\usepackage{url}
\usepackage{hyperref}
\usepackage{listings}
\usepackage[dvipsnames]{xcolor}
\usepackage{authblk}
\usepackage{amsmath}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}
\lstset{style=mystyle}

\hypersetup{%
  pdftitle={SMT-LIB Proposal: Integer Arithmetic with Exponentials},
  colorlinks=true,
  linkcolor=blue,
  citecolor=olive,
  filecolor=magenta,
  urlcolor=cyan
}

\bibliographystyle{plainurl}

\title{SMT-LIB Proposal: Integer Arithmetic with Exponentials}

\author[1]{Florian Frohn}
\author[2]{Hanna Lachnitt}
\author[1]{Philipp Schroer}
\author[3]{Yoni Zohar}
\affil[1]{RWTH Aachen University, Aachen, Germany}
\affil[2]{Stanford University, Stanford, USA}
\affil[3]{Bar-Ilan University, Ramat Gan, Israel}

\begin{document}

\maketitle

\section{Introduction}

We propose to extend the SMT-LIB theory {\tt Ints} with exponentiation.
%
Such an extension is motivated by various use cases, e.g.:\footnote{Further information on use cases and available benchmarks can be found here: \url{https://github.com/ffrohn/exp-smt-lib/issues/5}}
%
\begin{itemize}
\item SMT with parametric bit-vectors \cite{bitvec}
\item automatic verification of infinite state systems \cite{swine}
\item automatic verification of solutions for recurrence equations \cite{swine2}
\item deductive verification with Caesar \cite{caesar}
\end{itemize}

\section{Proposal}

The following proposal is based on the SMT-LIB theory {\tt Ints}.\footnote{\url{https://smt-lib.org/theories-Ints.shtml}}

\begin{lstlisting}[language=Lisp]
(theory Ints

 :sorts ((Int 0))

 :funs ((NUMERAL Int)
        (- Int Int)                   ; negation
        (- Int Int Int :left-assoc)   ; subtraction
        (+ Int Int Int :left-assoc)
        (* Int Int Int :left-assoc)
        (** Int Int Int)              ; exponentiation
        (div Int Int Int :left-assoc)
        (mod Int Int Int)
        (abs Int Int)
        (<= Int Int Bool :chainable)
        (<  Int Int Bool :chainable)
        (>= Int Int Bool :chainable)
        (>  Int Int Bool :chainable)
       )

 :funs-description
 "All ranked function symbols of the form
    ((_ divisible n) Int Bool)
  where n is a positive numeral.
 "

 :values
 "The set of values for the sort Int consists of
  - all numerals,
  - all terms of the form (- n) where n is a numeral other than 0.
 "

 :definition
 "For every expanded signature, the instance of Ints with that
  signature is the theory consisting of all Sigma-models that interpret:

  - the sort Int as the set of all integer numbers,

  - each numeral as the corresponding natural number,

  - (_ divisible n) as the function mapping to true all and only
    the integers that are divisible by n,

  - abs as the absolute value function,

  - div and mod according to Boute's Euclidean definition [1], that
    is, so as to satify the formula

    (forall ((m Int) (n Int))
      (=> (distinct n 0)
          (let ((q (div m n)) (r (mod m n)))
            (and (= m (+ (* n q) r))
                 (<= 0 r (- (abs n) 1))))))

  - ** as exponentiation if the second argument is non-negative.
    So in particular, (= (** 0 0) 1) holds.
    If n is negative, then ** is defined as to satisfy the formula

    (= (** m n) (div 1 (** m (- n))))

    So the following holds for all negative integers n:
    - (= (** 0 n) (div 1 0))
    - (= (** m n) (** m (- n))) if (= (abs m) 1)
    - (= (** m n) 0)            if (> (abs m) 1)

  - the other function symbols of Ints as expected.

  References:
  [1] Boute, Raymond T. (April 1992).
      The Euclidean definition of the functions div and mod.
      ACM Transactions on Programming Languages and Systems
      (TOPLAS) ACM Press. 14 (2): 127 - 144.
      doi:10.1145/128861.128862.
 "

 :notes
 "Regardless of sign of m,
  when n is positive, (div m n) is the floor of the rational number
  m/n; when n is negative, (div m n) is the ceiling of m/n.

  This contrasts with alternative but less robust definitions of
  div and mod where (div m n) is
  - always the integer part of m/n (rounding towards 0), or
  - always the floor of x/y (rounding towards -infinity).
 "

 :notes
 "See note in the Reals theory declaration about terms of the form
  (/ t 0). The same observation applies here to terms of the form
  (div t 0) and (mod t 0).
 "
)

\end{lstlisting}

\section{Justification}

While working on this proposal, several question have been discussed by the authors of this document, and may require further discussion in the community.

\subsection{Associativity}

Exponentiation\footnote{The details of the discussion are available here: \url{https://github.com/ffrohn/exp-smt-lib/issues/6}} is usually right-associative.
%
However, the authors agree that the use of exponentiation in an associative manner is far less common than for other associative operators like addition or multiplication.
%
Thus, declaring it as right-associative may cause bugs and confusion, and therefore keeping it strictly binary appears to be the better choice.

\subsection{Semantics for Negative Exponents}

We\footnote{The details of the discussion are available here: \url{https://github.com/ffrohn/exp-smt-lib/issues/2}} discussed several possibilities for the semantics of exponentiation with negative exponents.
%
We think that the following two variants are best:
%
\begin{itemize}
\item for all $n > 0$, we define $b^{-n} = 1 \mathrel{\mathsf{div}} b^n$, i.e.,
  \[
    b^{-n} =
    \begin{cases}
      1 \mathrel{\mathsf{div}} 0 & \text{if } \phantom{|}b\phantom{|} = 0\\
      b^n & \text{if } |b| = 1\\
      0 & \text{if } |b| > 1
    \end{cases}
  \]
  For the first case, the usual treatment of division by zero in SMT-LIB applies, i.e., the term $1 \mathrel{\mathsf{div}} 0$ is meaningful, but its interpretation is not restricted by the {\tt Ints} theory.
\item exponentiation with a negative exponent is treated like an uninterpreted function, as in the case of division by zero
\end{itemize}
%
The advantage of the first variant is that it avoids introducing further partial / uninterpreted functions.
%
The advantage of the second version is that it's least surprising from a user's perspective, in particular for users that are aware of the handling of division by zero in SMT-LIB.
%
After consulting with the SMT-LIB coordinators, we propose the first variant.

\subsection{Name of the Function Symbol for Exponentiation}

We\footnote{The details of the discussion are available here: \url{https://github.com/ffrohn/exp-smt-lib/issues/1}} considered several names, and identified {\tt **} and {\tt pow} as the best options.
%
In the proposal, we chose {\tt **}, as SMT-LIB usually uses ``operators'' (like {\tt +}, {\tt *}, etc.) instead of ``abbreviations'' if possible.
%
Note that we preferred {\tt **} over {\tt \^{}}, as the latter may be confused with bitwise XOR.

\subsection{Unary Exponential Functions}

We\footnote{The details of the discussion are available here: \url{https://github.com/ffrohn/exp-smt-lib/issues/3}} also considered including unary functions for exponentiation with fixed bases into the proposal:
%
\begin{lstlisting}[language=Lisp]
All ranked function symbols of the form
  ((n iexp _) Int Int)
where n is a positive numeral.
\end{lstlisting}
%
Here, \lstinline{iexp} abbreviates ``integer exponentiation''.
%
The motivation is that the extension of {\tt LIA} with one of these function symbols (which is also known as Semenov Arithmetic) is decidable, whereas the extension of {\tt LIA} with a binary function for exponentiation is undecidable.

We did not include such unary function symbols for simplicity.

\subsection{Logics}

{\color{BrickRed} If\footnote{The details of the discussion are available here: \url{https://github.com/ffrohn/exp-smt-lib/issues/4}} exponentiation gets incorporated into the {\tt Ints} theory, then all existing logics that refer to it have to be adapted in order to exclude exponentiation.}
%
Moreover, new logics that allow exponentiation should be defined.
%
We propose to define (at least) the following additional logics (names to be discussed):
%
\begin{itemize}
\item {\tt EIA}: {\tt NIA} + exponentiation
\item {\tt QF\_EIA}: {\tt QF\_NIA} + exponentiation
\item {\tt UFEIA}: {\tt UFNIA} + exponentiation
\end{itemize}
%
An extension of {\tt LIA} with exponentiation can simulate multiplication, so presumably, such a logic is not required.

\bibliography{refs}

\end{document}